The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2016-08-23 script/en/chapters/bitciphers.tex
CT Book ed 12: Major update -- many changes, corrections, and additions during last 4 months

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
% Pommerenings Specialities
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% \newcommand*{\C}{\mathbb{C}}   %BERM3
\renewcommand*{\C}{\mathbb{C}}   %BERM3
\newcommand*{\F}{\mathbb{F}}
\newcommand*{\M}{\mathbb{M}}
\renewcommand*{\N}{\mathbb{N}}   %BERM3
\renewcommand*{\Q}{\mathbb{Q}}   %BERM3
\renewcommand*{\R}{\mathbb{R}}   %BERM3
\renewcommand*{\Z}{\mathbb{Z}}   %BERM3
\newcommand*{\lsb}{\operatorname{lsb}}
\newcommand*{\Oh}{\operatorname{O}}

@@

\newpage
\hypertarget{Chapter_BitCiphers}{}
\chapter{Introduction to Bitblock and Bitstream Ciphers}
\label{Chapter_BitCiphers}
(Klaus Pommerening, 2015, Updates: Jan 2016, Apr 2016) \\

\noindent
While number theoretic methods prevail for the construction and analysis
of asymmetric encryption algorithms,
modern symmetric encryption algorithms\index{symmetric encryption} almost
always rely on Boolean algebra\index{Boolean algebra}\index{algebra!Boolean},
that is on the manipulation of bits. This involves a quite different kind of
Mathematics and might be unfamiliar to beginners. Therefore in this chapter
we attempt a smooth introduction into this mathematical subject. As previous
knowledge we assume elementary mathematical notions such as ``variable''
and ``function'', also for other domains than just the real numbers,
and knowledge of elementary algebra, calculus, and number theory.

@@

The focus of this chapter is on introducing the mathematical foundations
of ciphers that operate on bits. We won't define concrete ciphers in detail
but instead recommend the books by
Menezes/Orschot/Vanstone \cite{HAC},
Oppliger \cite{Opp},
Paar und Pelzl \cite{PaPe},
Schmeh \cite{Schm},
and Stamp \cite{Sta}.

@@

   {\bf Symmetric encryption of information given by bits}

The mathematical foundations and methods belong to the domains of

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|cc|ccc||cc|cc|} \hline
   \multicolumn{5}{|c||}{\bf logical} & \multicolumn{4}{c|}{\bf algebraic} \\ \hline
   \multicolumn{2}{|c|}{bits} & \multicolumn{3}{|c||}{composition} &
        \multicolumn{2}{|c|}{bits} & \multicolumn{2}{|c|}{composition} \\ \hline
   $x$ & $y$ & OR & AND & XOR & $x$ & $y$ & + & $\cdot$ \\ \hline
    F  &  F  & F  &  F  &  F  &  0  &  0  &  0  &  0    \\
    F  &  T  & T  &  F  &  T  &  0  &  1  &  1  &  0    \\
    T  &  F  & T  &  F  &  T  &  1  &  0  &  1  &  0    \\
    T  &  T  & T  &  T  &  F  &  1  &  1  &  0  &  1    \\
   \hline
\end{tabular}
\end{center}
\caption{The most important compositions of bits\index{bit}. The logical
  XOR\index{XOR} is identical with the algebraic +, the logical AND\index{AND},
  with the algebraic $\cdot$ (multiplication).}\label{t-bool-xor}
\end{table}

@@

For a comprehensive treatment see \cite{CuSt} or the two articles
by Claude Carlet\footnote{%
  also see his list of publications at
  \href{http://www.math.univ-paris13.fr/~carlet/pubs.html}{\tt http://www.math.univ-paris13.fr/$\sim$carlet/pubs.html}
} in \cite{CrHa}\footnote{%
  Carlet's articles are online at
  \href{http://www.math.univ-paris13.fr/~carlet/chap-fcts-Bool-corr.pdf}{\tt http://www.math.univ-paris13.fr/$\sim$carlet/chap-fcts-Bool-corr.pdf}
  and
  \href{http://www.math.univ-paris13.fr/~carlet/chap-vectorial-fcts-corr.pdf}{\tt http://www.math.univ-paris13.fr/$\sim$carlet/chap-vectorial-fcts-corr.pdf}
}.

As a first simple example, consider AND\index{AND} as a Boolean function:
It takes two bits and produces one new bit by the well-known rules
shown in Table~\ref{t-bool-xor}. For a slightly more complex example
take the function $f_0$ that produces the value
\begin{equation}\label{eq-expl-f0}
   f_0(x_1, x_2, x_3) = x_1\: \text{AND}\: (x_2\: \text{OR}\: x_3)
\end{equation}
from three bits $x_1$, $x_2$, $x_3$.

@@

\noindent The mechanism inside this black box may be specified from several different
points of view:
\begin{itemize}
   \item {\bf mathematically} by a formula
   \item {\bf informatically} by an algorithm
   \item {\bf technically} by a circuit\index{circuit}
      (or plugging diagram)
   \item {\bf pragmatically} by a truth table\index{truth table}
      (that is the complete lookup table of its values)
\end{itemize}
Our sample function $f_0$ is mathematically defined by Formula~(\ref{eq-expl-f0}).
The corresponding algorithm is also adequately specified by this formula since it
has no branching points or conditional statements. As a circuit we visualize $f_0$
as in Figure~\ref{fig-bool-circuit}. The truth table is in Table~\ref{tab-bool-wt}.

@@

The form of a Boolean function according to Theorem~\ref{thm-bool-disj}
is called  {\bf conjunctive normal
form\index{conjunctive normal form}\index{normal form!conjunctive}
(CNF\index{CNF})}. It is not unique\footnote{%
  For example we could add the terms
  $\wedge\: (x_1 \vee x_2) \wedge (x_1 \vee x_3)$ to the normal form of $f_0$.
}$^,$\footnote{%
  The SageMath class {\tt sage.logic.boolformula.BooleanFormula} provides
  transformations of a logical expression into CNF by the function
  {\tt convert\_cnf()}, and to the corresponding truth table by the
  function {\tt truthtable()}.
}.
Without further explanation we remark that there is a further simplification
as a ``canonical CNF'' that guarantees a certain uniqueness.
There is also an analoguous disjunctive normal
form\index{disjunctive normal form}\index{normal form!disjunctive}
({\bf DNF\index{DNF}}) (a ``disjunction of conjunctions'').

@@

Linear algebra over the field $\F_2$ is quite simple, many complications known
from other mathematical areas boil down to trivialities. Such is the
case for the the solution of systems of linear
equations\index{linear equation}\index{equation!linear},
explicitly written as
\[
\begin{matrix}
     a_{11} x_1 & + & \cdots & + & a_{1n} x_n & = & b_1 \\
     \vdots     &   &        &   & \vdots     &   & \vdots \\
     a_{m1} x_1 & + & \cdots & + & a_{mn} x_n & = & b_m
\end{matrix}
\]
with given $a_{ij}$ and $b_i \in \F_2$, and unknown $x_j$ for which we
search solutions. In matrix terms this system has an elegant expression as
\[
     A x = b
\]
where $A$ is an $m \times n$ matrix, and $x$ and $b$ are column vectors,
that is $n \times 1$ or $m \times 1$ matrices.

@@

In the general case (over an arbitrary field) the underlying algorithm for
solving a system of linear equations is Gaussian\footnote{%
  Johann Carl Friedrich Gauß\index{Gauss, Johann Carl Friedrich},
  German mathematician, astronomer, geodesist,
  and physicist, April 30, 1777 -- February 23, 1855
}
elimination. This algorithm of course also hides in the SageMath method {\tt solve\_right()}.

@@

The previous subsection described (and Table~\ref{tab-bool-3} illustrated)
how to interpret the bitblocks $x = (x_1, \ldots, x_n)$ of length $n$
as integers $i(x) = 0, 1, \ldots, 2^n-1$ in base-2 representation.
The example in Table~\ref{tab-bool-wt2} suggests how to describe the
truth table\index{truth table} of a Boolean function
$f\!\!: \F_2^n \longrightarrow \F_2$ in a parsimonious way by a bitblock
$b = (b_0, \ldots, b_{2^n-1})$ of length $2^n$:
simply take the last column in the order given by the indices $i(x)$.
The general procedure for arbitrary $n$ runs as follows:
\begin{align*}
     b_{i(x)} = f(x) \quad & \text{where }
     i(x) = x_1 \cdot 2^{n-1} + \cdots + x_{n-1} \cdot 2 + x_n \\
     & \text{for }\quad x = (x_1, \ldots, x_n) \in \F_2^n.
\end{align*}
This might look entangled, but it simply means: ``Interpret $x$ as
the base-2 representation of an integer $i(x)$, and set $f(x)$ as the
bit at position $i(x)$ from the bitblock $b$''. An additional column
$i(x)$ in the truth table of the function $f_0$, Table~\ref{tab-bool-wt},
illustrates this procedure, see Table~\ref{tab-bool-wt2}. The last
column of this t
able, written in row form, is the bitblock $b$.

In this way the bitblock $(0,0,0,0,0,1,1,1)$ completely specifies
the truth table of $f_0$ or, even more parsimoniously, this bitstring
of length $2^3 = 8$:
\[
     00000111
\]

@@

The algebraic normal form\index{algebraic normal form}\index{normal form!algebraic}
(ANF\index{ANF}) also has a characterization by $2^n$ bits: the
coefficients of the $2^n$ different monomials\footnote{%
  Remember that the ANF is a sum of monomials, and each monomial\index{monomial}
  is a product of a subset of $\{x_1, \ldots, x_n \}$, hence has a representation
  as an integer between $0$ und $2^n-1$.
} (see Theorem~\ref{thm-bool-anf1}, page~\pageref{thm-bool-anf1}). The
monomials also have an interpretation as bitblocks, see the list above.
Therefore we may view a bitblock $a = (a_0, \ldots, a_{2^n-1})$ as
representing the ANF of a Boolean function $f\!: \F_2^n \longrightarrow \F$
in the following way:
\begin{align*}
     f(x) = \sum_{i=0}^{2^n-1} a_i x_1^{e_1(i)} \cdots x_n^{e_n(i)} \quad & \text{where }
     i = e_1(i) \cdot 2^{n-1} + \cdots + e_n(i) \\
     & \text{with }\quad e_1(i), \ldots, e_n(i) = 0 \text{ or } 1.
\end{align*}
This formula means: ``Interpret the $n$-tuple $e$ of exponents of a
monomial as the base-2 representation of an integer $i$. The $i$-th
element of the bitblock $a$ indicates whether this monomial occurs in
the ANF of $f$ or not.''

@@

With these assumption we have a lower bound of about $80$ bits for
a secure key length according to the state of the art \cite{LeVe}.
Popular ciphers use keys of lengths $128$ or more, so have a sufficient
security margin. The outdated standard cipher DES used
56 bit keys. The technology of today breaks it quite quickly.

@@

\subsection{Algebraic
  Cryptanalysis}\label{ss-bool-algca}\index{algebraic cryptanalysis}\index{cryptanalysis!algebraic}

\subsubsection*{Attacks with Known\index{known plaintext}\index{plaintext!known}
  Plaintext\footnote{%
  An attack with known plaintext assumes that the attacker knows or
  guesses a small piece of plaintext and tries to deduce the key,
  or some more plaintext that is unknown to her. For the present
  section we assume that a complete plaintext bitblock is known.}}

@@

There are two powerful general approaches for solving systems of (polynomial)
equations over $\F_2$:
\begin{itemize}
   \item SAT solvers\index{SAT solver} \cite{GaJo}\footnote{%
      SAT denotes the satisfiability\index{satisfiability} problem
      of propositional logic. Consider a logical expression in Boolean variables
      $x_1, \ldots, x_n$ and ask if there exist values of the variables
      that make the expression ``True''. In other words consider a Boolean
      function\index{Boolean function}\index{function!Boolean} $f$ and ask
      if it assumes the value $1$. A {\bf SAT solver} is an algorithm that
      takes a logical expression\index{logical expression}\index{expression!logical}
      in CNF\index{CNF} and decides the satisfiability by finding a solution
      $x$, or showing there's no solution. The naive algorithm uses
      the truth table\index{truth table}
      and exhausts the $2^n$ possible arguments. However there are much faster
      algorithms, the most popular being the DPLL algorithm\index{DPLL algorithm}
      (after Davis, Putnam, Logemann, and Loveland) and BDD based\index{BDD}
      algorithms (Binary Decision Diagram). The SageMath modules {\tt sage.sat.solvers}
      and {\tt sage.sat.boolean\_polynomials} contain some of these algorithms.
      },
   \item elimination\index{elimination} using
      Groebner bases\index{Groebner basis} \cite{Bric}\footnote{%
      For an introduction to this theory see the textbooks \cite{Bard,CLOS,GaGe},
      or the script \cite{Sege}, or the paper \cite{Laza}.
      }.
\end{itemize}
Both methods work well for a small number of unknowns. With a growing number
of unknowns their complexity becomes unmanageable\footnote{%
  In fact SAT\index{SAT} was the first problem in history shown to be
  {\bf NP}-complete\index{NP-complete}.
}.
Of course we always find a solution by searching through the
complete value table. But this naive method is inefficient (exponential
in the number of unknowns, hopeless for 80 or more unknowns).
But also the costs of SAT solvers and Groebner-basis methods grow
exponentially with the number of unknowns. Not even the fact that
all partial degrees are $\leq 1$ is of vital help.

@@

\begin{theorem}[Garey/Johnson]
  The problem of finding a solution for a system of polynomial equations
  over $\F_2$ is {\bf NP}-complete\index{NP-complete}.
\end{theorem}
For a proof see the book by Garey/Johnson~\cite{GaJo}.\footnote{%
   A recent article on the difficulty of systems of polynom equations is
   \cite{CGHM}.
}

@@

To escape attacks bitblock ciphers\index{bitblock cipher}, or their round
maps, or their S-boxes\index{S-box}, should fulfill some requirements.
\begin{itemize}
\item {\bf Balance\index{balance}} All preimages have the same
   number of elements, or in other words, the values of the map are
   uniformly distributed. Irregularities of the distribution would
   provide hooks for statistical cryptanalysis.
\item {\bf Diffusion\index{diffusion}/avalanche effect\index{avalanche effect}}
   If a single plaintext bit changes, about 50\% of the ciphertext bits change.
   This effect conceals similarity of plaintexts.
\item {\bf Algebraic complexity}
   The determination of preimages or parts thereof should lead to equations
   whose solution is as difficult as possible. This requirement is related
   to the algebraic degree of the map, but only in an indirect way.
   A suitable measure is ``algebraic immunity''\index{algebraic immunity}\index{immunity!algebraic}.
\item {\bf Nonlinearity} We know several criteria that measure linearity,
   also ``hidden'' linearity, and are relatively easy to describe and to
   handle. For example they quantify how susceptible Boolean maps are for
   linear\index{linear cryptanalysis}\index{cryptanalysis!linear} or
   differential\index{differential cryptanalysis}\index{cryptanalysis!differential}
   cryptanalysis \cite{Pomm}.
   \begin{itemize}
   \item The linear potential\index{linear potential}\index{potential!linear}
      should be as low as possible, the linear\index{linear profile}\index{profile!linear}
      profile as balanced as possible.
   \item The differential\index{differential potential}\index{potential!differential}
      potential should be as low as possible, the differential
      profile\index{differential profile}\index{profile!differential} as balanced
      as possible.
    \end{itemize}
\end{itemize}
Some of these criteria are compatible with each other, some criteria contradict
other ones. Therefore the design of a bitblock cipher\index{bitblock cipher}
requires a balance between different criteria. Instead of optimizing a map
for a single criterion the designer should aim at a uniformly high level for
all criteria.

@@

This was easily done with pencil and paper. However it might be instructive
to retrace the count in SageMath for a better understanding of more complex examples.
SageMath sample~\ref{Sage-code-Mats-Expl} provides this. It uses the function
{\tt xor} from SageMath sample~\ref{Sage-code-bool-div-bbl} in Appendix~\ref{ss-bool-conv},
as well {\tt S0} from SageMath sample~\ref{Sage-code-Luc-S0}. The result
\mbox{\tt [2, 0, True]} says that we found $2$ zeroes among the counted values,
yielding the majority decision $0$, and the output parameter {\tt True} tells
that the decision was deterministic, not randomized.

@@

\subsection{Example A: A One-Round Cipher}\label{ss-bool-bsp1}

We consider examples that are much too simple for real
world applications but illustrate the principles of
linear\index{linear cryptanalysis}\index{cryptanalysis!linear}
cryptanalysis in an easily intelligible way. We always assume round
functions of the type $f(a+k)$, that is we add the key---or
an $n$-bit part of it---to the plaintext\footnote{%
   This is a quite special method of bringing the key into
   play but nevertheless realistic. The paradigmatic
   sample ciphers {\sc Lucifer}\index{Lucifer}, DES\index{DES},
   and AES\index{AES} do so. The term used with AES \cite{DaRi}
   is ``key-alternating cipher structure''.
} before applying a bijective
S-box $f\!: \F_2^n \longrightarrow \F_2^n$. The simplest model is
encryption by the formula
\[
   c = f(a+k),
\]
see\footnote{%
  The graphics here and later represent the map $f$ sometimes by the
  S-box S in the elementwise assignments.
} Figure~\ref{fig-bool-bsp0}.
This example is pointless because one block of known
plaintext\index{known plaintext}\index{plaintext!known} gives
a solution\footnote{%
   We assume that the attacker knows the inverse map $f^{-1}$ that
   is part of the decryption algorithm. One-way\index{one-way encryption}
   encryption methods that assume that $f^{-1}$ is not efficiently
   deducible from $f$ are the subject of another part of cryptography.
} for $k$:
\[
   k = f^{-1}(c) + a.
\]

@@

\subsection{Example B: A Two Round Cipher}\label{ss-bool-2rd}

@@

In computer science a feedback shift register is a special case of a finite-state
machine\index{finite-state machine}\index{machine!finite-state}.
Therefore the sequence of its states is periodic\index{periodic}.
Here is why.

>>>>

\begin{bibunit}[babalpha] %% alpha: Chapter bibliography shows authors abbreviation

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
% Pommerenings Specialities
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\newcommand*{\F}{\mathbb{F}}
\newcommand*{\M}{\mathbb{M}}

\newcommand*{\C}{\mathbb{C}}
\newcommand*{\N}{\mathbb{N}}
\newcommand*{\Q}{\mathbb{Q}}
\newcommand*{\R}{\mathbb{R}}
\newcommand*{\Z}{\mathbb{Z}}

\newcommand*{\lsb}{\operatorname{lsb}}
\newcommand*{\Oh}{\operatorname{O}}

@@

\newpage
\hypertarget{Chapter_BitCiphers}{}
\chapter{Introduction to Bitblock and Bitstream Ciphers}
\label{Chapter_BitCiphers}
(\hyperlink{author_Klaus-Pommerening}{Klaus Pommerening}, 2015, Updates: Jan 2016, Apr 2016) \\

\noindent
While number theoretic methods prevail for the construction and analysis
of asymmetric encryption algorithms,
modern symmetric encryption algorithms\index{encryption!symmetric} almost
always rely on Boolean algebra\index{Boolean algebra}\index{algebra!Boolean},
that is on the manipulation of bits. This involves a quite different kind of
Mathematics and might be unfamiliar to beginners. Therefore in this chapter
we attempt a smooth introduction into this mathematical subject. As previous
knowledge we assume elementary mathematical notions such as ``variable''
and ``function'', also for other domains than just the real numbers,
and knowledge of elementary algebra, calculus, and number theory.

@@

The focus of this chapter is on introducing the mathematical foundations
of ciphers that operate on bits. We won't define concrete ciphers in detail
but instead recommend the books by
Menezes/Orschot/Vanstone \cite{Menezes2001},
Oppliger \cite{Oppliger2011},
Paar und Pelzl \cite{PaPe2009},
Schmeh \cite{Schm2003, Schm2016},
and Stamp \cite{Stamp2007}.

@@

   {\bf Symmetric encryption of information given by bits}.

\noindent The mathematical foundations and methods belong to the domains of

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|cc|ccc||cc|cc|} \hline
   \multicolumn{5}{|c||}{\bf logical} & \multicolumn{4}{c|}{\bf algebraic} \\ \hline
   \multicolumn{2}{|c|}{bits} & \multicolumn{3}{|c||}{composition} &
        \multicolumn{2}{|c|}{bits} & \multicolumn{2}{|c|}{composition} \\ \hline
   $x$ & $y$ & OR & AND & XOR & $x$ & $y$ & + & $\cdot$ \\ \hline
    F  &  F  & F  &  F  &  F  &  0  &  0  &  0  &  0    \\
    F  &  T  & T  &  F  &  T  &  0  &  1  &  1  &  0    \\
    T  &  F  & T  &  F  &  T  &  1  &  0  &  1  &  0    \\
    T  &  T  & T  &  T  &  F  &  1  &  1  &  0  &  1    \\
   \hline
\end{tabular}
\end{center}
\caption{The most important compositions of bits\index{bit}. The logical
  XOR\index{XOR} is identical with the algebraic +, the logical AND\index{AND}
  with the algebraic $\cdot$ (multiplication).}\label{t-bool-xor}
\end{table}

@@

For a comprehensive treatment see \cite{CuSt2009} or \cite{Pomm2008, Pomm2014} or the two articles
by Claude Carlet\footnote{%
  also see his list of publications at
  \url{http://www.math.univ-paris13.fr/~carlet/pubs.html}
} in \cite{CrHa2010}\footnote{%
  Carlet's articles are online at
  \url{http://www.math.univ-paris13.fr/~carlet/chap-fcts-Bool-corr.pdf} and
  \url{http://www.math.univ-paris13.fr/~carlet/chap-vectorial-fcts-corr.pdf}
}.

As a first simple example, consider AND\index{AND} as a Boolean function:
It takes two bits and produces one new bit by the well-known rules
shown in Table~\ref{t-bool-xor}. For a slightly more complex example
take the function $f_0$ that produces the value
\begin{equation}\label{bc_sample-fct-f0-with-3-vars}%\label{eq-expl-f0}
   f_0(x_1, x_2, x_3) = x_1\: \text{AND}\: (x_2\: \text{OR}\: x_3)
\end{equation}
from three bits $x_1$, $x_2$, $x_3$.

@@

\noindent The mechanism inside this black box may be specified from several different
points of view:
\begin{itemize}
   \item {\bf mathematically} by a formula
   \item {\bf informatically} by an algorithm
   \item {\bf technically} by a circuit\index{circuit}
      (or plugging diagram)
   \item {\bf pragmatically} by a truth table\index{truth table}
      (that is the complete lookup table of its values)
\end{itemize}
Our sample function $f_0$ is mathematically defined by the
Formula~(\ref{bc_sample-fct-f0-with-3-vars}).
The corresponding algorithm is also adequately specified by this formula since it
has no branching points or conditional statements. As a circuit we visualize $f_0$
as in Figure~\ref{fig-bool-circuit}. The truth table is in Table~\ref{tab-bool-wt}.

@@

The form of a Boolean function according to Theorem~\ref{thm-bool-disj}
is called  {\bf conjunctive normal
form\index{conjunctive normal form}\index{normal form!conjunctive}
(CNF\index{CNF})}. It is not unique\footnote{%
  For example we could add the terms
  $\wedge\: (x_1 \vee x_2) \wedge (x_1 \vee x_3)$ to the normal form of $f_0$.
}$^,$\footnote{%
  The SageMath class {\tt sage.logic.boolformula.BooleanFormula} provides
  transformations of a logical expression to the CNF by the function
  {\tt convert\_cnf()}, and to the corresponding truth table by the
  function {\tt truthtable()}.
}.
Without further explanation we remark that there is a further simplification
as a ``canonical CNF'' that guarantees a certain uniqueness.
There is also an analoguous disjunctive normal
form\index{disjunctive normal form}\index{normal form!disjunctive}
({\bf DNF\index{DNF}}) (a ``disjunction of conjunctions'').

@@

Linear algebra over the field $\F_2$ is quite simple, many complications known
from other mathematical areas boil down to trivialities. Such is the
case for the solution of systems of linear
equations\index{linear equation}\index{equation!linear},
explicitly written as
\[
\begin{matrix}
     a_{11} x_1 & + & \cdots & + & a_{1n} x_n & = & b_1 \\
     \vdots     &   &        &   & \vdots     &   & \vdots \\
     a_{m1} x_1 & + & \cdots & + & a_{mn} x_n & = & b_m
\end{matrix}
\]
with given $a_{ij}$ and $b_i \in \F_2$, and unknown $x_j$ for which we
search solutions. In matrix terms this system has an elegant expression as
\[
     A x = b
\]
where $A$ is an $m \times n$ matrix, and $x$ and $b$ are column vectors,
that is $n \times 1$ or $m \times 1$ matrices.

@@

In the general case (over an arbitrary field) the underlying algorithm for
solving a system of linear equations is Gaussian\footnote{%
  Johann Carl Friedrich Gauß\index{Gauss, Carl Friedrich},
  German mathematician, astronomer, geodesist,
  and physicist, April 30, 1777 -- February 23, 1855
}
elimination. This algorithm of course also hides in the SageMath method {\tt solve\_right()}.

@@

The previous subsection described (and Table~\ref{tab-bool-3} illustrated)
how to interpret the bitblocks $x = (x_1, \ldots, x_n)$ of length $n$
as integers $i(x) = 0, 1, \ldots, 2^n-1$ in base-2 representation.
The example in Table~\ref{tab-bool-wt2} suggests how to describe the
truth table\index{truth table} of a Boolean function
$f\!\!: \F_2^n \longrightarrow \F_2$ in a parsimonious way by a bitblock
$b = (b_0, \ldots, b_{2^n-1})$ of length $2^n$:
simply take the last column in the order given by the indices $i(x)$.
The general procedure for arbitrary $n$ runs as follows:
\begin{align*}
     b_{i(x)} = f(x) \quad & \text{where }
     i(x) = x_1 \cdot 2^{n-1} + \cdots + x_{n-1} \cdot 2 + x_n \\
     & \text{for }\quad x = (x_1, \ldots, x_n) \in \F_2^n.
\end{align*}
This might look entangled, but it simply means: ``Interpret $x$ as
the base-2 representation of an integer $i(x)$, and set $f(x)$ as the
bit at position $i(x)$ from the bitblock $b$''. An additional column
$i(x)$ in the truth table of the function $f_0$
($f_0$ was defined in Formula~\ref{bc_sample-fct-f0-with-3-vars})
illustrates this procedure -- see Table~\ref{tab-bool-wt2}. The last
column of this table, written in row form, is the bitblock $b$.

In this way, the bitblock $(0,0,0,0,0,1,1,1)$ or, even more parsimoniously,
the bitstring
\[
     00000111
\]
of length $2^3 = 8$ completely specifies the truth table of $f_0$.\footnote{%
  In Python/SageMath bitblocks are implemented as lists\index{list}.
}

@@

The algebraic normal form\index{algebraic normal form}\index{normal form!algebraic}
(ANF\index{ANF}) also has a characterization by $2^n$ bits: the
coefficients of the $2^n$ different monomials\footnote{%
  Remember that the ANF is a sum of monomials. Each monomial\index{monomial}
  is a product of a subset of $\{x_1, \ldots, x_n \}$, and hence has a representation
  as an integer between $0$ und $2^n-1$.
}
(see Theorem~\ref{thm-bool-anf1}, page~\pageref{thm-bool-anf1}). The
monomials also have an interpretation as bitblocks, see the list above.
Therefore we may view a bitblock $a = (a_0, \ldots, a_{2^n-1})$ as
representation of the ANF of a Boolean function $f\!: \F_2^n \longrightarrow \F_2$
in the following way:
\begin{align*}
     f(x) = \sum_{i=0}^{2^n-1} a_i x_1^{e_1(i)} \cdots x_n^{e_n(i)} \quad & \text{where }
     i = e_1(i) \cdot 2^{n-1} + \cdots + e_n(i) \\
     & \text{with }\quad e_1(i), \ldots, e_n(i) = 0 \text{ or } 1.
\end{align*}
This formula means: ``Interpret the $n$-tuple $e$ of exponents of a
monomial as the base-2 representation of an integer $i$. The $i$-th
element of the bitblock $a$ indicates whether this monomial occurs in
the ANF of $f$ or not.''

@@

With these assumption we have a lower bound of about $80$ bits for
a secure key length according to the state of the art \cite{LeVe2000}.
Popular ciphers use keys of lengths $128$ or more, so have a sufficient
security margin. The outdated standard cipher DES used
56 bit keys. The technology of today breaks it quite quickly.

@@

\subsection{Algebraic Cryptanalysis}
\label{ss-bool-algca}\index{algebraic cryptanalysis}\index{cryptanalysis!algebraic}

\subsubsection*{Attacks with Known\index{known plaintext}\index{plaintext!known}
  Plaintext\footnote{%
  An attack with known plaintext assumes that the attacker knows or
  guesses a small piece of plaintext, and then tries to deduce the key
  or some more plaintext that is unknown to her. For the present
  section we assume that the known plaintext is a complete bitblock.}}

@@

There are two powerful general approaches for solving systems of (polynomial)
equations over $\F_2$:
\begin{itemize}
   \item SAT solvers\index{SAT solver} \cite{GaJo1979}\footnote{%
      SAT denotes the satisfiability\index{satisfiability} problem
      of propositional logic. Consider a logical expression in Boolean variables
      $x_1, \ldots, x_n$ and ask if there exist values of the variables
      that make the expression ``True''. In other words consider a Boolean
      function\index{Boolean function}\index{function!Boolean} $f$ and ask
      if it assumes the value $1$. A {\bf SAT solver} is an algorithm that
      takes a logical expression\index{logical expression}\index{expression!logical}
      in CNF\index{CNF} and decides the satisfiability by finding a solution
      $x$, or showing there's no solution. The naive algorithm uses
      the truth table\index{truth table}
      and exhausts the $2^n$ possible arguments. However there are much faster
      algorithms, the most popular being the DPLL algorithm\index{DPLL algorithm}
      (after Davis, Putnam, Logemann, and Loveland) and BDD based\index{BDD}
      algorithms (Binary Decision Diagram). The SageMath modules {\tt sage.sat.solvers}
      and {\tt sage.sat.boolean\_polynomials} contain some of these algorithms.
      },
   \item elimination\index{elimination} using
      Groebner bases\index{Groebner basis} \cite{Bric2010}\footnote{%
      For an introduction to this theory see the textbooks \cite{Bard2009,CLOS2007,GaGe1999},
      or the script \cite{Sege2004}, or the paper \cite{Laza1983}.
      }.
\end{itemize}
Both methods work well for a small number of unknowns. With a growing number
of unknowns their complexity becomes unmanageable\footnote{%
  In fact SAT\index{SAT} was the first problem in history shown to be
  {\bf NP}-complete\index{NP-complete}.
}.
Of course we always find a solution by searching through the
complete value table. But this naive method is inefficient (exponential
in the number of unknowns, and so hopeless for 80 or more unknowns).
But also the costs of SAT solvers and Groebner-basis methods grow
exponentially with the number of unknowns. Not even the fact that
all partial degrees are $\leq 1$ is of vital help.

@@

\begin{theorem}[Garey/Johnson]
  The problem of finding a solution for a system of polynomial equations
  over $\F_2$ is {\bf NP} complete\index{NP-complete}.
\end{theorem}
For a proof see the book by Garey/Johnson~\cite{GaJo1979}.\footnote{%
   A recent article on the difficulty of systems of polynom equations is
   \cite{CGHM2003}.
}

@@

To escape attacks bitblock ciphers\index{bitblock cipher}, or their round
maps, or their S-boxes\index{S-box}, should fulfill some requirements.
\begin{itemize}
\item {\bf Balance\index{balance}:} All preimages have the same
   number of elements, or in other words, the values of the map are
   uniformly distributed. Irregularities of the distribution would
   provide hooks for statistical cryptanalysis.
\item {\bf Diffusion\index{diffusion}/avalanche effect\index{avalanche effect}:}
   If a single plaintext bit changes, about 50\% of the ciphertext bits change.
   This effect conceals similarity of plaintexts.
\item {\bf Algebraic complexity:}
   The determination of preimages or parts thereof should lead to equations
   whose solution is as difficult as possible. This requirement is related
   to the algebraic degree of the map, but only in an indirect way.
   A suitable measure is ``algebraic immunity''\index{algebraic immunity}\index{immunity!algebraic}.
\item {\bf Nonlinearity:} We know several criteria that measure linearity,
   also ``hidden'' linearity, and are relatively easy to describe and to
   handle. For example they quantify how susceptible Boolean maps are for
   linear\index{linear cryptanalysis}\index{cryptanalysis!linear} or
   differential\index{differential cryptanalysis}\index{cryptanalysis!differential}
   cryptanalysis \cite{Pomm2014}.
   \begin{itemize}
   \item The linear potential\index{linear potential}\index{potential!linear}
      should be as low as possible, the linear\index{linear profile}\index{profile!linear}
      profile as balanced as possible.
   \item The differential\index{differential potential}\index{potential!differential}
      potential should be as low as possible, the differential
      profile\index{differential profile}\index{profile!differential} as balanced
      as possible.
    \end{itemize}
\end{itemize}
Some of these criteria are compatible with each other, some criteria contradict
other ones. Therefore the design of a bitblock cipher\index{bitblock cipher}
requires a balance between different criteria. Instead of optimizing a map
for a single criterion the designer should aim at a uniformly high level for
all criteria.

@@

This was easily done with pencil and paper. However it might be instructive
to retrace the count in SageMath for a better understanding of more complex examples.
SageMath sample~\ref{Sage-code-Mats-Expl} provides this. It uses the function
{\tt xor}\index{XOR} from SageMath sample~\ref{Sage-code-bool-div-bbl} in Appendix~\ref{ss-bool-conv},
as well {\tt S0} from SageMath sample~\ref{Sage-code-Luc-S0}. The result
\mbox{\tt [2, 0, True]} says that we found $2$ zeroes among the counted values,
yielding the majority decision $0$, and the output parameter {\tt True} tells
that the decision was deterministic, not randomized.

@@

\subsection{Example A: A one-round cipher}\label{ss-bool-bsp1}

We consider examples that are much too simple for real
world applications but illustrate the principles of
linear\index{linear cryptanalysis}\index{cryptanalysis!linear}
cryptanalysis in an easily intelligible way. We always assume round
functions of the type $f(a+k)$, that is we add the key---or
an $n$-bit part of it---to the plaintext\footnote{%
   This is a quite special method of bringing the key into
   play but nevertheless realistic. The paradigmatic
   sample ciphers {\sc Lucifer}\index{Lucifer}, DES\index{DES},
   and AES\index{AES} do so. The term used with AES \cite{DaRi2002}
   is ``key-alternating cipher structure''.
} before applying a bijective
S-box $f\!: \F_2^n \longrightarrow \F_2^n$. The simplest model is
encryption by the formula
\[
   c = f(a+k),
\]
see\footnote{%
  The graphics here and later represent the map $f$ sometimes by the
  S-box S in the elementwise assignments.
} Figure~\ref{fig-bool-bsp0}.
This example is pointless because one block of known
plaintext\index{known plaintext}\index{plaintext!known} gives
a solution\footnote{%
   We assume that the attacker knows the inverse map $f^{-1}$ that
   is part of the decryption algorithm. One-way\index{one-way encryption}
   encryption methods that assume that $f^{-1}$ is not efficiently
   deducible from $f$ are the subject of another part of cryptography.
} for $k$:
\[
   k = f^{-1}(c) + a.
\]

@@

\subsection{Example B: A two-round cipher}\label{ss-bool-2rd}

@@

In computer science a feedback shift register is a special case of a finite-state
machine\index{finite-state machine}\index{machine!finite-state}.
Therefore, the sequence of its states is periodic\index{period}.
Here is why.
